Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From The Reionization Epoch

### 2014-10-18 foreground_delay_spectrum_modeling.tex
From 071aa02: ============
(*) Changed color scale of figure 6 showing compact foreground model
(*) Made changes based on most of Jonnie's comments
(*) Major changes include justification of compact source flux density threshold
(*) Modified a couple of figures as per Jonnie's suggestion

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact objects dominate the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space.  Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate a large fraction of foreground contamination in redshifted 21~cm power spectrum analysis.

@@

A key challenge in the statistical detection of the redshifted H{\sc i} 21~cm signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}.

Considerable effort has been and continues to be made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,liu14a,liu14b}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

Foregrounds contaminate the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analyses. Here we use existing catalogs and a high fidelity instrumental model to capture both foreground and instrumental chromaticity. To decouple these effects from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual baseline--based approach of \citet{pob13} and \citet{par14}. This approximates the power spectrum as the inverse Fourier transform of the spectra generated by the instrument's correlator.

@@

Interferometer array data known as {\it visibilities}, $V_{bf}(\vec{b},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\vec{b}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. If $I(\hat{s},f)$ and $A(\hat{s},f)$ are the sky brightness and antenna's directional power--pattern, respectively, at different frequencies on the sky as a function of direction on the sky denoted by the unit vector ($\hat{s}$), and $W_f(f)$ denotes instrumental bandpass weights, then $V_{bf}(\vec{b},f)$ can be written as (with a slight adaptation from \citet{van34}, \citet{zer38}, and \citet{tho01}):
\begin{align}\label{eqn:obsvis}
  V_{bf}(\vec{b},f) &= \iint\limits_\textrm{sky} A(\hat{s},f)\,I(\hat{s},f)\,W_f(f)\,e^{-i2\pi f\frac{\vec{b}\cdot\hat{s}}{c}}\,\dif\Omega,
\end{align}
where, $c$ is the speed of light, and $\dif\Omega$ is the solid angle element along $\hat{s}$.

@@

Equation~\ref{eqn:obsvis} can be modified to express the Fourier decomposition of transverse sky brightness distribution as:
\begin{align}\label{eqn:vis}
  V_{uf}(\vec{u},f) &= \iint\limits_\textrm{sky} A(\hat{s},f)\,I(\hat{s},f)\,W_f(f)\,e^{-i2\pi\vec{u}\cdot\hat{s}}\,\dif\Omega,
\end{align}
where, $\vec{u}$ denotes the transverse spatial frequency vector. $\vec{u}$ is directly related to $\vec{b}$ as $\vec{u}\equiv \vec{b}/\lambda$, and to the transverse spatial wavenumber mode, $\vec{k}_\perp$, as $\vec{k}_\perp\equiv 2\pi\vec{u}/D(z)$, where, $D(z)$ is the transverse comoving distance at redshift $z$. Hence, $V_{uf}(\vec{u},f)\equiv V_{bf}(\lambda\vec{u},f)$. It is important to note that $\vec{b}$ is independent of frequency while $\vec{u}$ is not.

@@

While $V_{b\tau}(\vec{b},\tau)$ and $V_{u\eta}(\vec{u},\eta)$ in equations~\ref{eqn:obsvis} and \ref{eqn:vis} resemble each other very closely, they are not quite identical. The difference basically arises from the fact noted above that $\vec{b}$ is independent of frequency while $\vec{u}$ is not. \citet{par12} and \citet{liu14a} have discussed the mathematical correspondence between the two quantities.

@@

Note that since {\it visibility} is a complex quantity, its Fourier transform is Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna spacing vector. In this paper, $\vec{b}$ is assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the telescope site. Hence, a perfectly eastward oriented antenna spacing will observe an object in the eastern sky at positive delay, and conversely for an object in the western sky. Similarly, an object in the northern sky will appear at a positive delay at an antenna spacing oriented northward.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\vec{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off12,off10}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey (MWACS; Hurley-Walker et al. 2014, submitted). MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning; it has a flux density limit of 25~mJy and a declination range of -12\arcdeg~ to -40\arcdeg~ evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm -- based on forward modeling software by \citet{sul12} -- computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are fairly low signal--to--noise so we iteratively project along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antennas to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variations in cable delay. Finally, we fit for an additional phase known to be caused by reflections in 150~m length cables. %possibly could add elaboration, if I knew exactly which version of the cal was done here. doesn't really matter

@@

The instrument model consists primarily of the tile beam power--pattern, $A(\hat{s},f)$ (see equation~\ref{eqn:obsvis}). It is modeled as a mutually--coupled 4--by--4 dipole array with the overall power--pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14}. To speed up simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}. The power--patterns are shown in Figure~\ref{fig:power_pattern} for the pointings chosen in our study.

@@

@@

Based on these criteria, we selected 133 objects from the SUMSS catalog and 336 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission.

@@

@@

The Galactic center in the {\it off--zenith} pointing is one such example already available in our study. Figure~\ref{fig:sky-model} shows the sky model ({\it top:} compact objects, {\it bottom:} diffuse sky model) in this pointing. The Galactic center is the most dominant source of foreground contamination from the diffuse sky model and is co--located with a bright secondary lobe of the power pattern near the western horizon. Figure~\ref{fig:delay-map} shows the sky mapped to delays registered by the baseline vectors, of length 100~m for instance, oriented northward (top panel) and eastward (bottom panel). Since the Galactic center is located in the western sky, this figure demonstrates that it is observed at negative delay on a baseline vector oriented eastward and at zero delay on a baseline vector oriented northward. Figure~\ref{fig:baseline-breakup} shows the delay spectra on baselines oriented northward (67\fdg 5~$\le \theta_\textrm{b} <$~112\fdg 5) at the top and eastward (-22\fdg 5~$\le \theta_\textrm{b} <$~22\fdg 5) at the bottom. The Galactic center manifests itself most distinctly near the negative horizon delay limit on short eastward baselines in the delay spectrum (bottom panel of Figure~\ref{fig:baseline-breakup}). Consequently, the spillover caused by the instrument's transfer function from the {\it foreground wedge} into the {\it EoR window} is most severe in eastward baselines relative to northward baselines evident by the bright vertical stripes of foreground contamination.

\begin{figure*}[htb]
\centering
\subfloat[][Sky model]{\label{fig:sky-model}\includegraphics[width=0.33\linewidth]{figures/v1_0/sky_model_snapshot_0.eps}}
\subfloat[][Delay maps]{\label{fig:delay-map}\includegraphics[width=0.33\linewidth]{figures/v1_0/directional_delay_map_snapshot_0.eps}}
\subfloat[][Delay Spectra]{\label{fig:baseline-breakup}\includegraphics[width=0.33\linewidth]{figures/v1_0/baseline_orientation_breakup_snapshot_0.eps}}
\caption{(a): Sky model showing compact objects ({\it top}) and diffuse emission ({\it bottom}). Contours of power pattern are overlaid whose levels are identical to those in Figures~\ref{fig:DSM} and \ref{fig:CSM}. The Galactic center is very prominent in diffuse emission in the western sky and is co--located with a bright secondary lobe of the power pattern. (b): Sky hemisphere mapped to delays observed on antenna spacings with northward ({\it top}) and eastward ({\it bottom}) orientations. Color scale shown is for a 100~m antenna spacing. Delays vary linearly with antenna spacing length. The bright Galactic center in the western sky is recorded at zero delays in northward antenna spacings and close to negative horizon delay limit on eastward antenna spacings. (c): Simulated delay spectra $|V_{b\tau}(\vec{b},\tau)|$ on antenna spacings oriented northward ({\it top}) and eastward ({\it bottom}). The bright Galactic center is prominently visible close to negative horizon delay limit on eastward antenna spacings. These are also the most severely contaminated by foreground spillover. The northward antenna spacings, on the other hand, are the least contaminated.}
\label{fig:breakup}
\end{figure*}

@@

In our example, we discard eastward antenna spacings (-22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5) of lengths $|\vec{b}| < 40$~m. Figure~\ref{fig:before-after} shows the delay spectra obtained with all antenna spacings (top panel) and after applying our screening technique (bottom panel) on the {\it off--zenith} observation. Notice the remarkable reduction --- by about an order of magnitude --- in foreground contamination in the {\it EoR window} after application of our screening technique. The color scale range has been adjusted relative to the previous figures to show this significant reduction in foreground contamination.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/baseline_screening_before_after_nside_64_Tsys_95.0K.eps}
\caption{Simulated delay spectra amplitudes $|V_{b\tau}(\vec{b},\tau)|$ for the {\it off--zenith} pointing with all antenna spacings included ({\it top}) and with short eastward antenna spacings discarded ({\it bottom}). Discarded antenna spacings (black vertical stripes) have lengths $|\vec{b}| < 40$~m and orientations -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5. The spillover from the bright Galactic center near the negative horizon delay limit from the {\it foreground wedge} prominent when all antenna spacings are used is lowered by an order of magnitude when short eastward antenna spacings are discarded. Color scale has been adjusted to emphasize mitigation achieved in foreground contamination. \label{fig:before-after}}
\end{figure}

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination\footnote{We use a statistical measure of foreground contamination using the standard deviation of noiseless $V_{b\tau}(\vec{b},\tau)$ from foregrounds in the {\it EoR window}.} in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\vec{b}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7\fdg 5) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22\fdg 5). It is notable that short antenna spacings play maximum role in determining foreground contamination. For instance, among antenna spacings with orientations $|\theta_\textrm{b}|\le$~22\fdg 5, discarding those with lengths $|\vec{b}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity. On the other hand, a bright compact object at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or all of the eastward oriented antenna spacings.

>>>>

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact objects dominate the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space.  Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate foreground contamination substantially (up to a factor $\sim 10$ with negligible loss of data) in redshifted 21~cm power spectrum analysis.

@@

A key challenge in the statistical detection of the redshifted H{\sc i} 21~cm signal, via the spatial power spectrum of temperature fluctuations, arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}.

Considerable effort has been and continues to be made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,liu14a,liu14b}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide--field (and chromatic) response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

Foregrounds that have intrinsic spectral variations, instruments with high chromaticity, or approximations in power spectrum analyses can contaminate the true EoR H{\sc i} power spectrum. Here we use existing catalogs and a high fidelity instrumental model to capture both foreground and instrumental chromaticity. To decouple these effects from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual baseline--based approach of \citet{pob13} and \citet{par14}. This approximates the power spectrum as the inverse Fourier transform of the spectra generated by the instrument's correlator.

@@

Interferometer array data known as {\it visibilities}, $V_{bf}(\vec{b},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\vec{b}$ and then Fourier transformed along the time axis to obtain a spectrum along the frequency ($f$) axis. If $I(\hat{s},f)$ and $A(\hat{s},f)$ are the sky brightness and antenna's directional power--pattern, respectively, at different frequencies on the sky as a function of direction on the sky denoted by the unit vector ($\hat{s}$), and $W_f(f)$ denotes instrumental bandpass weights, then $V_{bf}(\vec{b},f)$ can be written as (with a slight adaptation from \citet{van34}, \citet{zer38}, and \citet{tho01}):
\begin{align}\label{eqn:obsvis}
  V_{bf}(\vec{b},f) &= \iint\limits_\textrm{sky} A(\hat{s},f)\,I(\hat{s},f)\,W_f(f)\,e^{-i2\pi f\frac{\vec{b}\cdot\hat{s}}{c}}\,\dif\Omega,
\end{align}
where, $c$ is the speed of light, and $\dif\Omega$ is the solid angle element along $\hat{s}$. We wish to emphasize that this equation is valid in general and does not involve any approximation.

@@

Equation~\ref{eqn:obsvis} can be modified to express the Fourier decomposition of transverse sky brightness distribution as:
\begin{align}\label{eqn:vis}
  V_{uf}(\vec{u},f) &= \iint\limits_\textrm{sky} A(\hat{s},f)\,I(\hat{s},f)\,W_f(f)\,e^{-i2\pi\vec{u}\cdot\hat{s}}\,\dif\Omega,
\end{align}
where, $\vec{u}$ denotes the transverse spatial frequency vector. $\vec{u}$ is directly related to $\vec{b}$ as $\vec{u}\equiv \vec{b}/\lambda$, and to the transverse spatial wavenumber mode, $\vec{k}_\perp$, as $\vec{k}_\perp\equiv 2\pi\vec{u}/D(z)$, where, $D(z)$ is the transverse comoving distance at redshift $z$. Hence, $V_{uf}(\vec{u},f)\equiv V_{uf}(\vec{b}/\lambda,f)\equiv V_{bf}(\lambda\vec{u},f)$. It is important to note that $\vec{b}$ is independent of frequency while $\vec{u}$ is not.

@@

While $V_{b\tau}(\vec{b},\tau)$ and $V_{u\eta}(\vec{u},\eta)$ in equations~\ref{eqn:obsvis} and \ref{eqn:vis} resemble each other very closely, they are not quite identical. The difference basically arises from the fact noted above that $\vec{b}$ is independent of frequency while $\vec{u}$ is not. \citet{par12} and \citet{liu14a} have discussed the mathematical correspondence between the two.

@@

Note that since {\it visibility} is a complex quantity, its Fourier transform is Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna spacing vector. In this paper, $\vec{b}$ is assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the telescope site. Hence, a perfectly eastward oriented antenna spacing will observe objects in the eastern and the western skies at positive and negative delays, respectively. Similarly, an object in the northern sky will appear at a positive delay at an antenna spacing oriented northward.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Fourier space in which delay (and power) spectra of EoR H{\sc i} signals are presented. The $x$--axis is denoted by $|\vec{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). Here, $k_\perp$ and $k_\parallel$ are obtained for a frequency of 185~MHz. The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions excluding the {\it wedge} and narrow extensions of the {\it wedge} (white unshaded strips) caused by convolution with the instrumental transfer function are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off12,off10}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey (MWACS; Hurley-Walker et al. 2014, submitted). MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning; it has a flux density limit of 25~mJy and a declination range of -12\arcdeg~ to -40\arcdeg~ evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm -- based on forward modeling software by \citet{sul12} -- computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are fairly low signal--to--noise so we iteratively project along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antennas to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variations in cable delay. Finally, we fit for an additional phase known to be caused by reflections in a few 150~m length cables. %possibly could add elaboration, if I knew exactly which version of the cal was done here. doesn't really matter

@@

The instrument model consists primarily of the tile beam power--pattern, $A(\hat{s},f)$ (see equation~\ref{eqn:obsvis}). It is modeled as a mutually--coupled 4--by--4 dipole array with the overall power--pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14}. To speed up simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}. Besides having the effect of adding a time--dependent uncertainty in the power pattern, these random delay fluctuations reduce the coherence in the phased addition of dipole signals resulting in deviations from predicted models of the power--pattern, most prominently at its nulls. The power--patterns so obtained are shown in Figure~\ref{fig:power_pattern} for the pointings chosen in our study.

@@

The model described in the preceding section is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, it is known to be devoid of bright compact sources of emission \citep{deo08}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and use compact sources above this threshold. It may be noted that slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0\fdg 85, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting compact sources which might be already blended in the diffuse sky model.

@@

Based on these criteria, we selected 133 objects from the SUMSS catalog and 336 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission. Figure~\ref{fig:CSM} shows the compact foreground emission model used in our study whose with contours of the power pattern overlaid. The flux densities of the compact sources are indicated by the color scale.

@@

It is worth emphasizing that the ideal foreground model is one with an angular resolution finer than that of the instrument that covers the whole sky at the observing frequency, a point we will revisit later for another reason. For this study, we note that our conclusions are not significantly affected despite minor differences in the thresholds from different radio source confusion estimates.

@@

The Galactic center in the {\it off--zenith} pointing is one such example already available in our study. Figure~\ref{fig:sky-model} shows the sky model ({\it top:} compact objects, {\it bottom:} diffuse sky model) in this pointing. The Galactic center is the most dominant source of foreground contamination from the diffuse sky model and is co--located with a bright secondary lobe of the power pattern near the western horizon. Figure~\ref{fig:delay-map} shows the sky mapped to delays registered by the baseline vectors, of length 100~m for instance, oriented northward (top panel) and eastward (bottom panel). Since the Galactic center is located in the western sky, this figure demonstrates that it is observed at negative delay on a baseline vector oriented eastward and at zero delay on a baseline vector oriented northward. Figure~\ref{fig:baseline-breakup} shows the delay spectra on baselines oriented northward (67\fdg 5~$\le \theta_\textrm{b} <$~112\fdg 5) at the top and eastward (-22\fdg 5~$\le \theta_\textrm{b} <$~22\fdg 5) at the bottom. The Galactic center manifests itself most distinctly near the negative horizon delay limit on short eastward baselines in the delay spectrum (bottom panel of Figure~\ref{fig:baseline-breakup}). Consequently, the spillover caused by the instrument's transfer function from the {\it foreground wedge} into the {\it EoR window} is most severe in eastward baselines (particularly the short ones) relative to northward baselines evident by the bright vertical stripes of foreground contamination.

\begin{figure*}[htb]
\centering
\subfloat[][Sky model]{\label{fig:sky-model}\includegraphics[width=0.33\linewidth]{figures/v1_0/sky_model_snapshot_0.eps}}
\subfloat[][Delay maps]{\label{fig:delay-map}\includegraphics[width=0.33\linewidth]{figures/v1_0/directional_delay_map_snapshot_0.eps}}
\subfloat[][Delay Spectra]{\label{fig:baseline-breakup}\includegraphics[width=0.33\linewidth]{figures/v1_0/baseline_orientation_breakup_snapshot_0.eps}}
\caption{(a): Sky model showing compact objects ({\it top}) and diffuse emission ({\it bottom}). Contours of power pattern are overlaid whose levels are identical to those in Figures~\ref{fig:DSM} and \ref{fig:CSM}. The Galactic center is very prominent in diffuse emission in the western sky and is co--located with a bright secondary lobe of the power pattern. (b): Sky hemisphere mapped to delays observed on antenna spacings with northward ({\it top}) and eastward ({\it bottom}) orientations. Color scale shown is for a 100~m antenna spacing. Delays vary linearly with antenna spacing length. The bright Galactic center in the western sky is recorded at zero delays in northward antenna spacings and close to negative horizon delay limit on eastward antenna spacings. (c): Simulated delay spectra $|V_{b\tau}(\vec{b},\tau)|$ on antenna spacings oriented northward ({\it top}) and eastward ({\it bottom}). The bright Galactic center is prominently visible close to negative horizon delay limit, especially on short eastward antenna spacings. These are also the most severely contaminated by foreground spillover. The northward antenna spacings, on the other hand, are the least contaminated.}
\label{fig:breakup}
\end{figure*}

@@

In our example, we discard eastward antenna spacings (-22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5) of lengths $|\vec{b}| < 30$~m. Figure~\ref{fig:before-after} shows the delay spectra obtained with all antenna spacings (top panel) and after applying our screening technique (bottom panel) on the {\it off--zenith} observation. Notice the remarkable reduction in foreground spillover into the {\it EoR window} via the removal of bright vertical stripes on short eastward antenna spacings resulting from the application of our screening technique. The color scale range has been adjusted relative to the previous figures to show this significant reduction in foreground contamination.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/baseline_screening_before_after_nside_64_Tsys_95.0K.eps}
\caption{Simulated delay spectra amplitudes $|V_{b\tau}(\vec{b},\tau)|$ for the {\it off--zenith} pointing with all antenna spacings included ({\it top}) and with short eastward antenna spacings discarded ({\it bottom}). Discarded antenna spacings (black vertical stripes) have lengths $|\vec{b}| < 30$~m and orientations -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5. The spillover from the bright Galactic center near the negative horizon delay limit from the {\it foreground wedge} prominent when all antenna spacings are used is lowered by an order of magnitude when short eastward antenna spacings are discarded. Color scale has been adjusted to emphasize mitigation achieved in foreground contamination. \label{fig:before-after}}
\end{figure}

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination\footnote{We use standard deviation of noiseless $V_{b\tau}(\vec{b},\tau)$ from foregrounds in the {\it EoR window} as a measure of foreground contamination.} in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\vec{b}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. The fraction of antenna spacings discarded relative to the total number of antenna spacings is shown in dotted lines for different ranges of $\theta_\textrm{b}$. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7\fdg 5) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22\fdg 5). It is notable that short antenna spacings play maximum role in determining foreground contamination. For instance, among antenna spacings with orientations $|\theta_\textrm{b}|\le$~22\fdg 5, discarding those with lengths $|\vec{b}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity as the fraction of discarded baselines increases from $\sim 5$\% to $\sim 25$\%. On the other hand, a bright compact object at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or, in the worst case, all of the eastward oriented antenna spacings accounting for $\sim 25$\% loss of data.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/baseline_screening_nside_64_Tsys_95.0K.eps}
\caption{Foreground contamination, estimated using standard deviation in the {\it EoR window}, for the {\it off--zenith} pointing as a function of orientation and length of discarded antenna spacings. Antenna spacings oriented eastward with varying degrees of directedness are considered: -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy the above criteria, ones with lengths $\le |\vec{b}|_\textrm{max}$ ($x$--axis) are discarded (dotted lines). Lower (dashed line) and upper (dot--dashed) limits on foreground contamination are derived from northward (67\fdg 5~$\le\theta_\textrm{b}<$~112\fdg 5) and eastward ($|\theta_\textrm{b}|\le$~22\fdg 5) antenna spacings respectively. Foreground contamination in the {\it EoR window} (solid lines) reduces by factor $\sim$1.5 ($|\theta_\textrm{b}|\le$~7\fdg 5) to $\sim$10 ($|\theta_\textrm{b}|\le$~22\fdg 5). Short baselines have maximum effect on foreground contamination. For instance, discarding antenna spacings oriented $|\theta_\textrm{b}|\le$~22\fdg 5 with lengths $|\vec{b}|\gtrsim 30$~m has no effect in further reducing foreground contamination. \label{fig:screening}}
\end{figure}
